The \emph{FrobeniusThresholds} package for \emph{Macaulay2}

### 2019-06-06 Fsing/FThresholdsPaper/FThresholds.tex
Changes to Section 2 of the paper

\documentclass{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,setspace}
\RequirePackage[dvipsnames,usenames]{xcolor}

@@

It turns out that the sequence $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ itself, and not just its limit, also encodes interesting information about $f$.  For example, it is closely related to the so-called \emph{Bernstein-Sato polynomial} \cite{MustataTakagiWatanabeFThresholdsAndBernsteinSato}.  Thus, it is natural to ask whether one can recover the sequence $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ from the limit $\ft^{\idealm}(f)$.  Fortunately, the answer to this question is yes \cite{MustataTakagiWatanabeFThresholdsAndBernsteinSato, HernandezFPurityOfHypersurfaces}. More precisely, 
\begin{equation}
\label{sequence-from-limit}
\nu_f^{\idealm}(p^e) = \lceil p^e \cdot \ft^{\idealm}(f) \rceil - 1.  
\end{equation}

@@

We first consider the {\tt nu} function, a fundamental component of the package \emph{FrobeniusThresholds}.   As in the introduction, in this section we work in a polynomial ring over a finite field of characteristic $p>0$.

@@

%R = ZZ/11[x,y];
%I = ideal(x^2 + y^3, x*y);
%J = ideal(x^2, y^3);
%nu(1, I, J)
%f = x*y*(x^2 + y^2);
%nu(1, f, J)

@@

If the third argument is omitted from the function {\tt nu}, the third argument is assumed to be the homogeneous maximal ideal of the polynomial ring.

@@

i11 : M = ideal(x, y, z);

@@

i12 : nu(2, f) == nu(2, f, M)

@@

As pointed out in the introduction, if $\mathfrak{m}$ is the homogeneous maximal ideal and $f \in \mathfrak{m}$, then for all $e$, the values $\nu^{\mathfrak{m}}_f(p^e)$ can be recovered from the $F$-pure threshold of $f$.  
This fact is used to speed up computations for certain polynomials whose $F$-pure thresholds can be computed quickly via specialized algorithms, namely diagonal polynomials, binomials, and forms in two variables (see Section \ref{sec.FPT}).  This feature can be disabled by setting the option {\tt UseSpecialAlgorithms} (default value {\tt true}) to {\tt false}.  Consider the following example, where $f$ is a diagonal polynomial.

@@

In general, the function {\tt nu} works by searching through the list of integers $n$, and checking containments of the $n^\text{th}$ power of $I$ in a specified Frobenius power of $J$.  
The user can specify how this search is approached through the  option {\tt Search}.  Namely, a binary search is used when {\tt Search} is set to {\tt Binary} (and this is the option's default value), and a linear search is run when it is set to {\tt Linear}.

@@

If the option {\tt ReturnList} is changed from its default value of {\tt false} to {\tt true}, instead of simply outputting $\nu_I^J(p^e)$, {\tt nu} outputs a list of the values $\nu_I^J(p^s)$, for $s=0,\ldots,e$.
Alternatively, the same information can be found by setting the option {\tt Verbose} to {\tt true} (its default value is {\tt false}), to request that each of these values $\nu_I^J(p^s)$ is printed as they are iteratively computed.  This can also serve a way to monitor the progress of the computation.

@@

i9 : nu(5, f, Verbose => true)
nu(1) = 0
nu(p^1) = 1
nu(p^2) = 8
nu(p^3) = 44
nu(p^4) = 224
nu(p^5) = 1124
\end{MyVerbatim}
}

@@

Finally, we describe the option {\tt ContainmentTest} for the function {\tt nu}.
This option can be set to {\tt FrobeniusPower}, {\tt FrobeniusRoot}, or {\tt StandardPower}. The default value of this option depends on what is passed to {\tt nu}. Indeed, by default, {\tt ContainmentTest} is set to {\tt FrobeniusRoot} if the second argument passed to {\tt nu} is a polynomial, and is set to {\tt StandardPower} if it is an ideal.  We describe the consequences of setting {\tt ContainmentTest} to each of these values below, using an ideal $I$ as the second argument of $\nu$, since the option works analogously when $I$ is replaced with a polynomial $f$.

First, if {\tt ContainmentTest} is set to {\tt StandardPower}, then the ideal containments checked when computing {\tt nu(e,I,J)} are verified directly.
That is, a standard power of $I$ is first computed, say $I^n$, and a check is then run to see if the resulting ideal lies in the $p^e$-th Frobenius power of $J$.

%{\color{red} \[ I \subseteq J^{[p^e]} \iff I \subseteq I^{[1/p^e]} \subseteq J.\]}

Alternately, if {\tt ContainmentTest} is set to {\tt FrobeniusRoot}, then the ideal containments are verified through the computation of Frobenius roots.  That is, the $p^e$-th Frobenius root of $I^n$ is first computed, and a check is then run to see if it lies in $J$.  The output is unaffected, but this option often speeds up computations, especially when a polynomial or principal ideal is passed as the second argument.

%R = ZZ/11[x,y,z];
%f = x^3 + y^3 + z^3 + x*y*z;
%time nu(3, f) -- ContainmentTest is set to FrobeniusRoot, by default
%time nu(3, f, ContainmentTest => StandardPower)

Finally, suppose that {\tt ContainmentTest} is set to {\tt FrobeniusPower}.  Instead of producing the invariant $\nu_I^J(p^e)$, in this case the {\tt nu} function instead outputs the maximal integer $n$ such that the $n^{\textrm{th}}$ \emph{(generalized) Frobenius power} of $I$ is not contained in the $p^e$-th Frobenius power of $J$ \cite{hernandez+etal.frobenius_powers}. 
Note that the resulting value coincides with $\nu_I^J(p^e)$ when $I$ is replaced with a polynomial, or is a principal ideal, but need not coincide in general. 
For example, the two commands below give different output values. 
{\small
\setstretch{.67}
\begin{MyVerbatim}

@@

o4 = 8
\end{MyVerbatim}
}
\medspace
\noindent In line with the philosophy described in the introduction, after appropriately normalizing and taking the limit of these values as $e$ approaches infinity, the values computed when {\tt ContainmentTest} is set to {\tt FrobeniusPower}
limit to certain invariants of interest called \emph{critical exponents} (see \cite{hernandez+etal.frobenius_powers}).

>>>>

\documentclass{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,amsfonts,setspace}
\RequirePackage[dvipsnames,usenames]{xcolor}

@@

It turns out that the sequence $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ itself, and not just its limit, also encodes interesting information about $f$.  For example, it is closely related to the so-called \emph{Bernstein-Sato polynomial} \cite{MustataTakagiWatanabeFThresholdsAndBernsteinSato}.  Thus, it is natural to ask whether one can recover the sequence $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ from the limit $\ft^{\idealm}(f)$.  Fortunately, the answer to this question is yes \cite{MustataTakagiWatanabeFThresholdsAndBernsteinSato, HernandezFPurityOfHypersurfaces}. More precisely, 
\begin{equation*}
\nu_f^{\idealm}(p^e) = \lceil p^e \cdot \ft^{\idealm}(f) \rceil - 1.  
\end{equation*}

@@

We first consider the {\tt nu} function, a fundamental component of the package \emph{FrobeniusThresholds}.   As in the introduction, in this section we work in a polynomial ring over a finite field of characteristic $p>0$.  As in the introduction, $\idealm$ will denote the ideal of this ring generated by the variables.

@@

% R = ZZ/11[x,y];
% I = ideal(x^2 + y^3, x*y);
% J = ideal(x^2, y^3);
% nu(1, I, J)
% f = x*y*(x^2 + y^2);
% nu(1, f, J)

@@

If the third argument is omitted from the function {\tt nu}, it is assumed to be the maximal ideal $\idealm$.

@@

i11 : m = ideal(x, y, z);

@@

i12 : nu(2, f) == nu(2, f, m)

@@

As pointed out in the introduction, if $f \in \mathfrak{m}$, the values $\nu^{\mathfrak{m}}_f(p^e)$ can be recovered from the $F$-pure threshold of $f$.  
This fact is used to speed up computations for certain polynomials whose $F$-pure thresholds can be computed quickly via specialized algorithms, namely diagonal polynomials, binomials, and forms in two variables (see Section \ref{sec.FPT}).  This feature can be disabled by setting the option {\tt UseSpecialAlgorithms} (default value {\tt true}) to {\tt false}.  Consider the following example, where $f$ is a diagonal polynomial.

@@

In general, the function {\tt nu} works by searching through the list of integers $n$, and checking containments of the $n$-th power of $I$ in a specified Frobenius power of $J$.  
The user can specify how this search is approached through the  option {\tt Search}.  Namely, a binary search is used when {\tt Search} is set to {\tt Binary} (and this is the option's default value), and a linear search is run when it is set to {\tt Linear}.

@@

If the option {\tt ReturnList} is changed from its default value of {\tt false} to {\tt true}, rather than simply outputting $\nu_I^J(p^e)$, {\tt nu} outputs a list of the values $\nu_I^J(p^s)$, for $s=0,\ldots,e$.
Alternatively, the same information can be found by setting the option {\tt Verbose} to {\tt true} (its default value is {\tt false}), to request that each of these values $\nu_I^J(p^s)$ is printed as they are iteratively computed.  This can also serve a way to monitor the progress of the computation.

@@

i9 : nu(5, f, Verbose => true)
nu(1) = 0
nu(p^1) = 1
nu(p^2) = 8
nu(p^3) = 44
nu(p^4) = 224
nu(p^5) = 1124
\end{MyVerbatim}
}
\medspace

@@

As described in the introduction, the integer $\nu_I^J(p^e)$ is defined in terms of the containment conditions on powers of ideals. 
For positive integers $n$ and $e$, and an ideal $I$ contained in the radical of another ideal $J$, we have that 
\begin{equation*}
  I^n \subseteq J^{[p^e]} \Longleftrightarrow (I^n)^{[1/p^e]} \subseteq J 
\end{equation*}
where $(I^n)^{[1/p^e]}$ denotes the $p^e$-th \emph{Frobenius root} of $I^n$, as defined in \cite{BlickleMustataSmithDiscretenessAndRationalityOfFThresholds}.
The option {\tt ContainmentTest} for {\tt nu} allows the user to choose which of the two types of containment statements appearing above to use toward the calculation of $\nu_I^J(p^e)$.

If {\tt ContainmentTest} is set to {\tt StandardPower}, then {\tt nu(e,I,J)} is computed by testing the left-hand containment above, and when it is set to {\tt FrobeniusRoot}, the right-hand containment is checked. 
For efficiency reasons, the default value for {\tt ContainmentTest} is set to  {\tt FrobeniusRoot} if the second argument passed to {\tt nu} is a polynomial, and is set to {\tt StandardPower} if it is an ideal.

i1 : R = ZZ/11[x,y,z];

i2 : f = x^3 + y^3 + z^3 + x*y*z;

i3 : time nu(3, f) -- ContainmentTest is set to FrobeniusRoot, by default
     -- used 0.116697 seconds

o3 = 1209

i4 : time nu(3, f, ContainmentTest => StandardPower)
     -- used 9.65013 seconds

o4 = 1209
\end{MyVerbatim}
}
\medspace

% R = ZZ/11[x,y,z];
% f = x^3 + y^3 + z^3 + x*y*z;
% time nu(3, f) -- ContainmentTest is set to FrobeniusRoot, by default
% time nu(3, f, ContainmentTest => StandardPower)

The option {\tt ContainmentTest} has a third possible value, {\tt FrobeniusPower}, which may produce a different result than the other two values.  
Given an ideal $I$, the first, third, and fourth authors introduced the notion of a (generalized) Frobenius power $I^{[n]}$ of $I$, when $n$ is an arbitrary positive integer, not necessarily of the form $n = p^e$ \cite{hernandez+etal.frobenius_powers}.    
When {\tt ContainmentTest} is set to {\tt FrobeniusPower}, rather than  computing $\nu_I^J(p^e)$, the function {\tt nu} computes the 
maximal integer $n$ for which $I^{[n]}$ is not contained in $J^{[p^e]}$.  This number is denoted $\mu_I^J(p^e)$ in \emph{loc.\ cit.}, and equals $\nu_I^J(p^e)$ when $I$ is replaced with a polynomial, or is a principal ideal, but need not coincide in general. 
For example, the two commands below give different output values.

@@

o4 = 8
\end{MyVerbatim}
}

\medspace
\noindent In line with the philosophy described in the introduction, after appropriately normalizing and taking the limit as $e$ approaches infinity, the values $\mu_I^J(p^e)$ determine certain invariants of interest called \emph{critical exponents}.
